Modelling fine-scale correlation structures using hidden Markov models

### 2020-07-17 Report/Sections/intro.tex

Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales. In this work, we consider an example where killer whales exhibit a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale. One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017} \citep{Adam:2019}. HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

All HMMs presented so far assume Markovian dynamics in the underlying process (i.e. that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps). At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.  With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.

One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution to a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. Moreover, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to perform inference.

For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}. In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017}, specifically fluking \citep{Shorter:2017} from accelerometer data. Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.

We consider several models: the classical hierarchical hidden Markov model (HHMM) \citep{Barajas:2017}, the CarHMM model of \citep{Lawler:2019}, and our new model that blends the two (CarHHMM). The HHMM consists of a coarse-scale process and a fine-scale process, with standard HMMs modelling both the coarse- and fine-scale processes. The CarHMM modifies the usual HMM by modelling additional dependence into the sequence of observed data. In our combination of the HHMM and the CarHHMM models, we keep the usual Markov structure for the coarse-scale process, but use the CarHMM model for the fine-scale process.

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. We then perform a simulation study to compare the performance each existing model with ours in a controlled setting. Finally, we apply our method to dive data collected from a Northern resident killer whale off the coast of British Columbia, Canada.

>>>>

Another issue that arises in high-frequency data is that several simultaneous behavioral processes may occur at different time scales. In this work, we consider an example where killer whales exhibit a variety of different types of dives at a coarse scale, but also exhibit many different types of swimming behaviors within dives at a fine scale. One solution to this issue is to use a hierarchical hidden Markov model (HHMM) \citep{Barajas:2017,Adam:2019}. HHMMs model the entire time series in question as a nested structure of hidden Markov models (HMM) where each HMM corresponds to one behavioral process.

HMMs assume Markovian dynamics in the underlying process (i.e. that any observation $Y_t$ depends only on the behavioral state $X_t$ and $Y_{t-1}$ when conditioned on all previous time steps). At the shortest time scales, however, observations often exhibit complicated dependence structures which cannot be easily captured by traditional HMMs, CarHMMs, or HHMMs. Examples included periodic fluking behavior in killer whales off the coast of Vancouver, BC, and swimming patterns of horn sharks of the coast of Southern California \citep{Adam:2019}.  With improvements in tagging technology allowing for data to be collected at very high frequencies, noisy and non-Markovian fine-scale processes are likely to persist.

One solution is to model the fine-scale behavior using a continuous-time model, which involves modelling the dynamics of an animal as the solution of a stochastic differential equation. Continuous-time models are more flexible than their discrete-time counterparts and can incorporate observations taken at irregular time intervals. Unfortunately, they are often computationally intractable and require approximate inference techniques such as Markov-chain Monte Carlo (MCMC) methods to fit.

For periodic behavior in particular, one way to avoid the use of continuous time models is to use signal processing techniques such as the Fourier transform on the raw data. The advantages of using Fourier analysis within an HMM has been recently demonstrated in the context of describing daily behavioral cycles of marine mammals \citep{Heerah:2017}. In addition, Fourier analysis has previously been used in the field of animal movement to explain animal behavior \citep{Fehlmann:2017} and specifically fluking \citep{Shorter:2017} from accelerometer data. Thus, incorporating Fourier analysis of accelerometer data within the structure of an HMM appeared a promising simple approach to account for additional correlation in data that is cyclical in nature.

This work investigates how to incorporate fine-scale processes into the larger structure of hierarchical hidden Markov models while maintaining computational efficiency. We describe a general procedure that can be used to extract features from highly structured fine-scale behaviors that otherwise could not be modeled with existing HMM models. In addition, we bridge the gap between the discrete CarHMM and certain continuous-time stochastic process models by showing that the two are equivalent under certain conditions. We then perform a simulation study to compare the performance each existing model with ours in a controlled setting. Finally, we apply our method to dive data of a killer whale (\textit{Orcinus orca}) from the threatened Northern resident population off the coast of British Columbia, Canada.

### 2020-07-17 Report/Sections/results.tex

The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the behavioral state must be taken into account when using accelerometer data \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales. This, in turn, can assist conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. A plot of the raw data for all dives and a collection of five selected dives can be seen in Figure \ref{fig:data} and Figure \ref{fig:data_one_dive}, respectively.

@@

The acceleration exhibits significant sinusoidal behavior at several points in time (see fig \ref{fig:data_one_dive}), so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window. Acceleration data was a 3-dimensional vector, so $Z^*$ was calculated as follows:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2$$
%
$\mathbf{z}_{t,t^*}^{*(1)}$ is a 3-dimensional vector while $z_t^{*(2)}$ is a scalar. Summing the first 10 Fourier modes to calculate $z^{(2)}_{t,t^*}$ corresponds to a maximum frequency of $\tilde \omega = 5$ Hz. Figure \ref{fig:lag} displays a lag plot for all major features and reveals strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions. Auto-correlation was therefore directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
%
$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
%
where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.

While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear. Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$. The distribution of $Z^{*(2)}_{t,t^*}$ was also assumed to be gamma:
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2$$
%
$Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent when conditioned of the sub-dive state $X_{t,t^*}$.

It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically and admittedly somewhat arbitrarily. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is nearly identical to the one from the simulation study, with the exception that the fine scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a vector rather than a scalar. The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHMM-DFT. The coarse-scale model is comprised of an HMM with no auto-correlation and no DFT with hidden states corresponding to dive types and observations corresponding to dive durations. The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window). Refer to Figure \ref{fig:CarHHMM} for a full graphical representation of this model.

@@

Table \ref{table:emis_dists} displays estimates of the emission distribution parameters. Each fitted distribution is also plotted in Figure \ref{fig:coarse_emis} and Figure \ref{fig:fine_emis}. While the ecological interpretation of behavioral states is tenuous for HMMs, we hypothesize the following meanings. Dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including as rest before dives of type 2, which are deeper and more sustained. Sub-dive behavioural state 1 corresponds to gliding and less overall activity compared to the other behavioral states. The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavior 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state. Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component are much higher than every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds. Finally, sub-dive state 2 corresponds to a moderate amount of activity, as almost every parameter estimate is between the other two behavioral states.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.849 & 0.151 \\
0.907 & 0.093
\end{pmatrix}$$
$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix}$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.724 & 0.276 & 0.000 \\
0.057 & 0.887 & 0.056 \\
0.000 & 0.247 & 0.753
\end{pmatrix} \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.871 & 0.129 & 0.000 \\
0.135 & 0.829 & 0.036 \\
0.000 & 0.246 & 0.754
\end{pmatrix}$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix} \qquad
\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
%
About 86\% of observed dives are short; the whale usually rests for many dives in a row before performing a deep dive. The probability transition matrix $\hat \Gamma$ shows that the distribution of a particular dive's type is approximately equal regardless of the previous dive type. The fine-scale probability transition matrices imply that the killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives- 14\% of shallow dives are spent in sub-dive state 1 while 48\% of deep dives are spent in sub-dive state 1. Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath- a phenomenon which has been observed in bottle-nose dolphins \citep{Williams:1999}. (Fig \ref{fig:labeled_dives}) shows the decoded dive behavior of 5 selected dives, and (fig \ref{fig:coarse_probs}) and (fig \ref{fig:fine_probs}) display the probability of each dive type and sub-dive state, respectively.

@@

Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual of a particular observation is the marginal CDF of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}. To easily visualize outliers, the pseudo-residual is often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ is $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified. $Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (fig \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.

In addition to psuedoresiduals, we plotted histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state. This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state. If the model is correct, then the histograms of features should closely resemble their respective fitted probability distribution. Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is right-skewed. In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3 (fig. \ref{fig:empirical_dist}), indicating the existence of rare events corresponding to very violent thrashing of the killer whale. These outliers are potential subjects for future study.

>>>>

The CarHHMM-DFT was used to analyze dive data from a Northern Resident Killer Whale (NRKW) off the coast of British Columbia, Canada. Acceleration data can be a good proxy for energy expenditure \citep{Green:2009}, but studies suggest that the animal's behavioral state must be taken into account to obtain accurate estimates \citep{Dot:2016}. Therefore, understanding both the behavioral state of the killer whale as well as the distribution of accelerometer data within each behavioral state is import to understand the energetic requirements of killer whales. This knowledge can help ecologists understand the animal's energetic requirements and which in turn can help conservation efforts.

@@

The data used in this study was collected on September 2, 2019 from 12:49 pm to 6:06 pm and consists of depth and acceleration in three orthogonal directions. Observations were collected at a rate of 50 Hz. Tagging the killer whale caused anomalous behavior before 1:20 pm and after 6:00 pm, so observations in this time range were ignored. In addition, the tagging technology dropped data between 2:25pm and 2:37pm as well as between 4:07 and 5:07 pm, so any partially observed data within this time range were ignored as well. A killer whale ``dive" is considered to be any continuous chunk of data that occurs below 0.5 meters in depth and lasts for at least 10 seconds. Accelerometer and depth data were smoothed by taking a moving average with a window of 1/10th of a second. Data preprocessing was done in part with the \textit{divebomb} package in Python \citep{Nunes:2018}. After preprocessing the raw data, a total of 267 dives were observed. Figure \ref{fig:data} displays the dive profile and accelerometer data.

@@

The acceleration exhibits significant sinusoidal behavior, so the fine-scale observations $Z^*$ were made up of the DFT summary statistics of a two-second sliding window. Unlike the simulation study, the acceleration observations here are 3-dimensional vectors rather than scalars, so the observations $z^*$ were calculated as follows:
%
$$\mathbf{z}_{t,t^*}^{*(1)} := \mathcal{R}\left(\hat{\mathbf{y}}^{*(0)}_{t,t^*}\right) \qquad z_{t,t^*}^{*(2)} := \frac{1}{100}\sum_{k=1}^{10}||\hat{\mathbf{y}}^{(k)}_{t,t^*}||^2.$$
%
The observations are therefore made up of $\mathbf{z}_{t,t^*}^{*(1)}$, a 3-dimensional vector, and $z_t^{*(2)}$, a scalar. We calculate $z_t^{*(2)}$ by summing the first 10 Fourier modes, and corresponds to a maximum recorded frequency of $\tilde \omega = 5$ Hz. There is strong auto-correlation within $\mathbf{Z}^{*(1)}_{t,t^*}$ for all dimensions (see the supplementary material for a lag plot). Therefore, auto-correlation was directly modeled into the the distribution of $\mathbf{Z}^{*(1)}_{t,t^*}$, which is assumed to be Normally distributed with the following parameters:
%
$$\mathbb{E}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = \mathbf{z}, X^*_{t,t^*} = i) = \phi_1^{*(i)} \mathbf{z} + (1-\phi_1^{*(i)}) \mathbf{\mu}_1^{*(i)}$$
$$\mathbb{V}(\mathbf{Z}^{*(1)}_{t,t^*}|\mathbf{Z}^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \text{diag}\left[\left(\mathbf{\sigma}_1^{*(i)}\right)^2\right]$$
%
where $\phi_1^{*(i)} \in \mathbb{R}$, $\mathbf{\mu}_1^{*(i)} \in \mathbb{R}^3$, and $\mathbf{\sigma}_1^{*(i)} \in \mathbb{R}^3$.

While $Z^{*(2)}_{t,t^*}$ also exhibits some auto-correlation, the relationship is less strong, and the biological interpretation of auto-correlation within $Z^{*(2)}_{t,t^*}$ is less clear. Auto-correlation was therefore not incorporated into the emission distribution of $Z^{*(2)}_{t,t^*}$. In particular, the distribution of $Z^{*(2)}_{t,t^*}$ was assumed to be gamma and parameterized by its mean and variance:
%
$$\mathbb{E}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \mu_2^{*(i)}$$
$$\mathbb{V}(Z^{*(2)}_{t,t^*}|Z^{*(1)}_{t,t^*-1} = z,X^*_{t,t^*} = i) = \left(\sigma_2^{*(i)}\right)^2.$$
%
The observations $Z^{*(2)}_{t,t^*}$ and $\mathbf{Z}^{*(1)}_{t,t^*}$ were assumed to be independent of one another when conditioned of the sub-dive state $X_{t,t^*}$.

It is known that information criteria tends to overestimate the number of states in biological processes \citep{Pohle:2017}, so we instead selected $N = 2$ dive types and $N^* = 3$ sub-dive behaviours heuristically. The absence of principled method to select the number of hidden states is a common issue in statistical ecology, so it is important to use model validation techniques in lieu of information criteria (see section \ref{subsec:model_validation}).

The final model is nearly identical to the one from the simulation study, with the exception that the fine-scale Markov chain has three sub-dive behaviors instead of two ($N^* = 3$), and that the observation $\mathbf{z}^{*(1)}_{t,t^*}$ is a 3-dimensional vector rather than a scalar. The model is comprised of two levels: a coarse-scale HMM and a fine-scale CarHMM-DFT. The coarse-scale model is comprised of an HMM with hidden states corresponding to dive types and observations corresponding to dive durations. The fine-scale model is comprised of a CarHMM-DFT where auto-correlation is modeled into the distribution of $\mathbf{Z}^{*(1)}$ (the average acceleration with a 2-second window), but not $Z^{*(2)}$ (the ``wiggliness" of the two second window). Refer to Figure \ref{fig:CarHHMM-DFT} for a full graphical representation of this model.

@@

Table \ref{table:emis_dists} displays estimates of the emission distribution parameters for this case study. The fitted emission distributions are also plotted in Figures \ref{fig:coarse_emis} and \ref{fig:fine_emis}. While the ecological meaning behind behavioral states is tenuous for HMMs, we hypothesize the following interpretations. Dive type 1 corresponds to shorter, shallower dives which serve a variety of purposes, including as rest before dives of type 2, which are deeper and more sustained. Sub-dive behavioural state 1 corresponds to gliding and less overall activity compared to the other behavioral states. The mean of $Z^{*(2)}$ in this state is at least an order of magnitude smaller than sub-dive behavioral state 2, the variance of $\mathbf{Z}^{*(1)}$ is smaller than sub-dive behavior 2 for every component, and the auto-correlation of $\mathbf{Z}^{*(1)}$ is higher than every other behavioral state. Sub-dive state 3, on the other hand, corresponds to vigorous swimming activity, as the mean of $Z^{*(2)}$ and variance of $\mathbf{Z}^{*(1)}$ for every component is much higher than every other state. The auto-correlation of $\mathbf{Z}^{*(1)}$ is also much lower in this state, implying more variation in acceleration every 2 seconds. Finally, sub-dive state 2 corresponds to a moderate amount of activity, as almost every parameter estimate is between the other two behavioral states.

The estimated probability transition matrices and associated stationary distributions are
%
$$\hat \Gamma = \begin{pmatrix} 
0.849  &  0.151 \\
0.907  &  0.093
\end{pmatrix},$$
$$\hat \delta = \begin{pmatrix} 0.857 & 0.143 \end{pmatrix},$$
%
$$\hat \Gamma^{*(1)} = \begin{pmatrix} 
0.724 & 0.276 & 0.000 \\
0.057 & 0.887 & 0.056 \\
0.000 & 0.247 & 0.753
\end{pmatrix}, \qquad 
\hat \Gamma^{*(2)} = \begin{pmatrix} 
0.871 & 0.129 & 0.000 \\
0.135 & 0.829 & 0.036 \\
0.000 & 0.246 & 0.754
\end{pmatrix},$$
$$\hat \delta^{*(1)} = \begin{pmatrix} 0.143 & 0.698 & 0.159 \end{pmatrix}, \qquad
\hat \delta^{*(2)} = \begin{pmatrix} 0.476 & 0.456 & 0.067 \end{pmatrix}.$$
%
About 86\% of observed dives are short; the whale usually rests for many dives in a row before performing a deep dive. The probability transition matrix $\hat \Gamma$ shows that the probability of a particular dive type does not depend much on the previous dive type. This killer whale is much more likely to be in a less active sub-dive state when performing deep dives than when performing shallow dives:  sub-dive state 1 makes up 14\% of shallow dives, but it makes up 48\% of deep dives. Using less active sub-dive states when diving deep could be an energy reduction strategy for these long periods of holding breath. This phenomenon has been observed in bottle-nose dolphins \citep{Williams:1999} and may also be the case for killer whales. Figure \ref{fig:labeled_dives} shows the decoded dive behavior of 5 selected dives. The supplementary material also shows the estimated \textit{probability} of each dive time and sub-dive hidden state.

@@

Two visual tools were used to evaluate this model: pseudo-residuals and empirical histograms. A pseudo-residual of a particular observation is the marginal CDF of an observation conditioned on all other observations under the learned model \citep{Zucchini:2016}. To easily visualize outliers, this pseudo-residual is often passed through the quantile function of the standard Normal distribution. Mathematically, the pseudo-residual of an observation $y_t$ in a traditional HMM is equal to $\Phi^{-1} \left(Pr(Y_t < y_t|\{Y_1,\ldots,Y_T\}/\{Y_t\}) \right)$, where $\Phi$ is the cumulative distribution function of a standard Normal distribution. If the model is correct, then all pseudo-residuals are independent and follow a standard Normal distribution. We find that histograms of the pseudoresiduals of this model mostly support that the model is well-specified. $Z^{*(2)}$ is an exception, as its pseudo-residuals are noticeably right-skewed (Figure \ref{fig:pseudoresids}). This implies that the true distribution of $Z^{*(2)}$ may follow a heavier-tailed distribution compared to a gamma distribution such as a power law.

We also plotted separate histograms of each feature where each observation was weighted by the probability that the whale was in a particular hidden state. This empirical distribution was then plotted over the fitted probability distribution function of that feature and hidden state. If the model is correct, then the histogram of each feature and hidden state should closely resemble its corresponding fitted probability distribution. See Figure \ref{fig:empirical_dist} for an example of the dive duration. Our results mostly support a well-specified model with the exception of $Z^{*(2)}$, which is again right-skewed. In addition, $\mathbf{Z}^{*(1)}$ has heavy tails for sub-dive state 3, indicating the existence of rare events corresponding to exceptionally violent thrashing of the killer whale. These outliers are potential subjects for future study. See the supplementary material for empirical distributions of every feature and every hidden state.

### 2020-07-17 Report/main.tex

Recent advances in high-frequency tagging
technology have made a vast
amount of movement data available in a variety of fields. This
rich data can exhibit simultaneous behavioural processes occurring at different
time scales, resulting in very complicated dependence structures. These processes can be modelled 
through a hierarchical hidden Markov model
(HHMM), where the system is modelled as a
nested structure of hidden Markov models (HMMs).
At very short time scales, however, many assumptions of traditional HMMs are violated. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from a northern resident killer whale off
the coast of British Columbia, Canada.

@@

%%%%%%%%%%%%
\vfill\eject
{\centerline{\Large{Notes from Nancy}}}
\input{Sections/0627NH}
\vfill\eject
%%%%%%%%%%%%%
%
\section{Introduction}
\input Sections/intro.tex
%
\section{Models and parameter estimation}
\input Sections/models.tex
%
\section{Simulation Study}
\input Sections/simulation.tex
%
\section{Killer Whale Case Study}
\label{sec:case_study}
\input Sections/results.tex
%
\section{Discussion}
\input Sections/discussion.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{Figures and Tables}
\input Sections/figures.tex
%
%\CJShistory

>>>>

Recent advances in high-frequency tagging
technology have made a vast
amount of movement data available in a variety of fields. This data can exhibit simultaneous behavioural processes occurring at different
time scales, resulting in increasingly complicated dependence structures. These processes can be modelled 
through a hierarchical hidden Markov model
(HHMM), which models the system as a
nested structure of hidden Markov models (HMMs).
At very short time scales, however, many basic assumptions of traditional HMMs are violated. We demonstrate how
to incorporate fine-scale processes into the
larger structure of HHMMs while maintaining computational efficiency. We apply our
method to dive and accelerometer data collected from a northern resident killer whale off
the coast of British Columbia, Canada.

@@

%%%%%%%%%%%%
%\vfill\eject
%{\centerline{\Large{Notes from Nancy}}}
%\input{Sections/0627NH}
%\vfill\eject
%%%%%%%%%%%%%
%
\section{Introduction}
\input Sections/intro.tex
%
\section{Models and parameter estimation}
\input Sections/models.tex
%
\section{Simulation Study}
\input Sections/simulation.tex
%
\section{Killer Whale Case Study}
\label{sec:case_study}
\input Sections/results.tex
%
\section{Discussion}
\input Sections/discussion.tex
%
\newpage
\bibliography{references}
%
\newpage
\begin{appendix}
\input Sections/appendix.tex
\end{appendix}
%
\newpage
\section{Figures and Tables}
\input Sections/figures.tex
%
%\CJShistory
